Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the `eechidna` `R` package.

### 2019-03-31 draft-paper.Rmd
added explanation of factors created + justification of doing single PCA, and edited general dimension reduction section

*1: HERE Introductory paragraph*
Electoral socio-demographic variables are now available for each of the six elections and are joined with their corresponding two-party preferred votes. Before choosing an appropriate model, two issues with the socio-demographic variables are addressed. Firstly, variable scales change over the years and as such the data is standardized accordingly. Secondly, many variables represent similar information and where appropriate, should be combined in some way. Principal component analysis is used to determine which intuitive groups of variables covary and should be combined into a single variable. This also reduces the dimension of the data. After these steps, a model specification is chosen.

@@

### Dimension reduction
With only $N = 150$ observations (electorates) in each election and $p = 65$ socio-demographic variables in each cross-section, any model using all variables would face serious problems with multi-collinearity and over-fitting, likely leading to erroneous conclusions regarding variable significance. Therefore a form of dimension reduction is adopted before any models are fitted.

Socio-demographic variables[^2] that represent similar information are combined into "factors", informed by principal component analysis (PCA) on a combined dataset of socio-demographic variables across the six elections. A factor is created by combining several variables that have large loadings in a particular component and an intuitive justification for their grouping. Only the first four principal components are considered (noting that the scree plot levels off after the fourth component). A loading with magnitude greater than 0.15 is considered large. After computing these *sums*, each factor is again standardized to have mean zero and variance one, within each election.

*2: HERE Explain the different factors and how they are constructed - also justify combining all cross-sections*

The scree plots of the principal components for each election all level off after four components, and the loadings of these four components are similar across the elections. Principal components are then computed on the combined set of socio-demographics across all six elections. A factor is created by combining several variables all have large loadings in a particular component and when there is an intuitive reason as to why these variables could represent common information. A loading with magnitude greater than 0.15 is considered large. After computing these sums, each factor is again standardized to have mean zero and variance one, within each election.

Consider the `Incomes` factor as an illustration. Independent of principal components, we may suspect that median personal income, median household income and median family income all provide similar information about the financial wellbeing of an electorate. Their loadings on the first principal component are large (0.19, 0.21 and 0.22 respectively), suggesting that these variables should be combined into a single factor.

Having produced these *Seven* factors, the predictor set is reduced to $p = 30$. All of these variables are used in the regression for each election.

>>>>

*1: HERE Introductory paragraph*
Electoral socio-demographic variables are now available for each of the six elections and are joined with their corresponding two-party preferred votes. Before choosing an appropriate model, two issues with the socio-demographic variables are addressed. Firstly, variable scales change over the years and as such the data is standardized accordingly. Secondly, many variables represent similar information and where appropriate, should be combined in some way. Principal component analysis is used to identify variables covary, from which intuitive groupings are selected to be combined into a single variable. This also reduces the dimension of the data. After these steps, a model specification is chosen.

@@

### Dimension reduction
There are only $N = 150$ observations (electorates) in each election and $p = 65$ socio-demographic variables in each cross-section, with many variables represent similar information about an electorate. Any model that uses all variables would face serious problems with multi-collinearity and over-fitting, which would likely lead to erroneous conclusions regarding variable significance. To address this, groups of variables that represent similar information are combined into "factors"[^2].

A factor is created from a group of variables if there is an intuitive reason as to why they should represent similar information and if there is evidence to suggest that they covary. For example, a potential group would be variables relating to electoral incomes - median family, household and personal incomes. To determine which variables covary, principal component analysis is used on a combined dataset of socio-demographic variables from all six elections. It is appropriate to compute principal components in this way because when computed separately for each election, scree plots level off after four components and the loadings of the first four components are similar across the elections.

*2: HERE - ADD JUSTIFICATION FOR DOING SINGLE PCA AND EXPLAIN FACTORS CREATED*

Only the first four principal components from the combined dataset are considered, as the scree plot levels off after the fourth component. Variables that have a large loading in a particular component are deemed to covary, with a loading with magnitude greater than 0.15 being considered large. Six factors are created using this criteria. These are; `Incomes` - median personal, household and family incomes, `Unemployment` - unemployment and labor force participation rates, `ProportyOwned` - rates of housing ownership, mortgages, renting and government housing, `RentLoanPrice` - median rental and loan repayments, `FamHouseSize` - average household size, ratio of people to families and household makeup (single person, couple with kids and couple without kids) and `Education` - high school and university qualifications, jobs requiring higher levels of education as well as vocational course completions and jobs that do not require higher education levels (e.g. laborer, tradesperson). For each of these groups, variables with positive loadings are added and those with negative loadings are substracted to create a factor. After computing these sums, each factor is standardized to have mean zero and variance one, within each election.

There are $p=30$ variables in the resultant predictor set, with all of these used in the regression for each election.
